\Large The state learner -- a super learner for right-censored data

### 2024-12-20 manuscript/manuscript.tex
rewordings in the introduction

We define the state learner, a new super learner for right-censored
data, which simultaneously estimates the expected loss of learners of
the event time distribution and the censoring distribution. The loss
function underlying the state learner operates on the coarsened
data. In the right-censored survival setting the coarsened data
consist of the minimum and the order of the censoring time and the
event time as well as the baseline covariates. The state learner can
include a broad class of survival models as learners, it can handle
competing risks, and it does not require a single pre-specified
estimator of the conditional censoring distribution. To analyse the
theoretical properties of the state learner we focus on the discrete
super learner which combines the library of learners by picking the
one that minimises the cross-validated loss \citep{van2007super}. In
the presence of competing risks, our algorithm uses separate libraries
of learners for the cumulative hazard functions for each of the
competing risks and for the censoring distribution. We show that the
oracle selector of the state learner is consistent if all libraries
contain a consistent learner and prove a finite sample oracle
inequality.

@@

There are many existing super learners for right-censored data. We briefly
review them here and contrast our suggestion. Machine learning based on
right-censored data commonly uses the partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with right-censored data
by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
time. However, the partial log-likelihood loss does not work well with data
splitting (cross-validation) in continuous time because this loss assigns an
infinite value when applied to a test set time point which does not occur in the
learning set as soon as the learner predicts piece-wise constant cumulative
hazard functions. This is the case for prominent survival learners including the
Kaplan-Meier estimator, the random survival forest, and the semi-parametric Cox
regression model, which hence cannot be included in the library of the super
learner proposed by \cite{polley2011-sl-cens}.

When a proportional hazards model is assumed, the baseline hazard function can
be profiled out of the likelihood \citep{cox1972regression}. The cross-validated
partial log-likelihood loss \citep{verweij1993cross} has therefore been
suggested as a loss function for super learning \citep{golmakani2020super}. This
choice of loss function restricts the library of learners to include only Cox
proportional hazards models, and hence excludes many learners such as, e.g.,
random survival models, additive hazards models and accelerated failure time
models.
% A lesser
% known fact is that a super learner constructed with the partial
% log-likelihood loss implicitly depends on the censoring distribution
% \citep{hjort1992inference,whitney2019comment}.

Other attempts at super learning with right-censored data use an inverse
probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
% An IPCW loss function is attractive because the associated risk does
% not depend on the censoring distribution but describes a feature of
% the population of interested. Similar results can be obtained using
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
All these methods rely on an estimator of the censoring distribution,
and their drawback is that this estimator has to be pre-specified.
% When the data-generating mechanism is complex and not
% well-understood, pre-specification of the censoring distribution is a
% challenge.
% Inverse probability of censoring weighted loss has been studied for
% decades \citep{graf1999assessment, hothorn2006survival,
  % gerds2006consistent}. A disadvantage of inverse probability of
% censoring weighted loss estimators is that they require a
% pre-specified model for the censoring distribution.
% Recently \cite{westling2021inference} proposed a procedure which
% iterates between super learning of the censoring distribution and
% super learning of the event time distribution.

A recent approach which avoids a pre-specified censoring model was proposed
independently by \cite{han2021inverse} and \cite{westling2021inference}. In both
articles, the authors suggest to iterate between learning of the outcome model
and learning of the censoring model using IPCW loss functions. However, no
general theoretical guarantees seem to exist for this procedure. For instance,
the procedure is not guaranteed to converge to the true data-generating
mechanism \citep{munch2024thesis}.

Contrary to the super learners considered by \cite{polley2011-sl-cens} and
\cite{golmakani2020super}, the state learner poses no restrictions on the type
of learners that can be included in the library. Contrary to IPCW, censoring
unbiased transformations, or pseudo-value based super learners, the state
learner does not require a pre-specified censoring distribution, but constructs
a super learner for both the outcome and the censoring distribution. Contrary to
the proposals by \cite{han2021inverse} and \cite{westling2021inference}, the
state learner is backed by theoretical guarantees.

>>>>

We define the state learner, a new super learner for right-censored
data, which simultaneously estimates the expected loss of learners of
the event time distribution and the censoring distribution. Compared
to the super learners of \cite{polley2011-sl-cens} and
\cite{golmakani2020super}, the state learner proposed here poses no
restrictions on the type of learners that can be included in the
library. Compared to super learners that rely on inverse probability
of censoring weighting, censoring unbiased transformations, or
pseudo-value based super learners, the state learner does not require
a pre-specified censoring distribution. Compared to the super learners
by \cite{han2021inverse} and \cite{westling2021inference}, the state
learner has theoretical guarantees which these methods do not
have.
% Moreover, state learner seems to be the first super learner for
% right censored data which learns the outcome and the censoring
% distribution simulaneously.

The loss function underlying the state learner operates on the
coarsened data. In the right-censored survival setting the coarsened
data consist of the minimum and the order of the censoring time and
the event time as well as the baseline covariates. The state learner
can include a broad class of survival models as learners, it can
handle competing risks, and it does not require a single pre-specified
estimator of the conditional censoring distribution. To analyse the
theoretical properties of the state learner we focus on the discrete
super learner which combines the library of learners by picking the
one that minimises the cross-validated loss \citep{van2007super}. In
the presence of competing risks, our algorithm uses separate libraries
of learners for the cumulative hazard functions for each of the
competing risks and for the censoring distribution. We show that the
oracle selector of the state learner is consistent if all libraries
contain a consistent learner and prove a finite sample oracle
inequality.

@@

There are many existing super learners for right-censored data. We
briefly review them here. Machine learning based on right-censored
data commonly uses the partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with
right-censored data by \cite{polley2011-sl-cens}, where data is
assumed to be observed in discrete time. However, the partial
log-likelihood loss does not work well with data splitting
(cross-validation) in continuous time. The reason is that the partial
log-likelihood loss assigns an infinite value for any learner that
predicts piece-wise constant cumulative hazard functions when there
are event times in the test set which do not occur in the learning
set. This problem occurs with prominent survival learners including
the Kaplan-Meier estimator, the random survival forest, and the
semi-parametric Cox regression model and these learners cannot be
included in the library of the super learner proposed by
\cite{polley2011-sl-cens}. When a proportional hazards model is
assumed, the baseline hazard function can be profiled out of the
likelihood \citep{cox1972regression}. The cross-validated partial
log-likelihood loss \citep{verweij1993cross} has therefore been
suggested as a loss function for super learning
\citep{golmakani2020super}. This choice of loss function restricts the
library of learners to include only Cox proportional hazards models,
and hence excludes many learners such as, e.g., random survival
models, additive hazards models and accelerated failure time models.
Other proposals for super learning with right-censored data use an
inverse probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
All these methods rely on an estimator of the censoring distribution,
and their drawback is that this estimator has to be pre-specified. An
approach which avoids a pre-specified censoring model was proposed
independently by \cite{han2021inverse} and
\cite{westling2021inference}. In both articles, the authors suggest to
iterate between learning of the outcome model and learning of the
censoring model using IPCW loss functions. However, no general
theoretical guarantees seem to exist for this procedure and the
procedure is not guaranteed to converge to the true data-generating
mechanism \citep{munch2024thesis}.
